---
title: "Exercise 4.5: Bayesian Inference for Poisson/Gamma Model"
author: "Rajitha M. Silva"
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this exercise, we analyze a dataset where observed counts follow a **Poisson/Gamma** distribution. The **Poisson/Gamma model** is commonly used to model count data with overdispersion (i.e., variance greater than the mean).

We estimate the **parameters** \(a\) and \(b\) in the Poisson/Gamma model using **Bayesian inference**.

# Problem Setup

We assume the data \( y_1, ..., y_n \) follows a **Poisson/Gamma** density:

\[
f(y|a, b) = \frac{\Gamma(y + a)}{\Gamma(a) y!} \cdot \frac{b^a}{(b + 1)^{y + a}}
\]

where:
- \( a > 0 \) is the **shape parameter**,
- \( b > 0 \) is the **rate parameter**.

We assume a **noninformative prior**:

\[
p(a, b) \propto \frac{1}{(ab)^2}
\]

To simplify computations, we transform parameters:

\[
\theta_1 = \log a, \quad \theta_2 = \log b.
\]

Then, the posterior density is:

\[
p(\theta_1, \theta_2 | y) \propto \frac{1}{ab} \prod_{i=1}^{n} \frac{\Gamma(y_i + a)}{\Gamma(a) y_i!} \cdot \frac{b^a}{(b + 1)^{y_i + a}}
\]

where \( a = e^{\theta_1} \) and \( b = e^{\theta_2} \).

# Observed Data

We use **data from Gilchrist (1984)**, where **33 insect traps** were set across sand dunes, recording the number of insects of the **Staphylinoidea** taxa caught:

```{r}
counts <- c(2, 5, 0, 2, 3, 1, 3, 4, 3, 0, 3,
            2, 1, 1, 0, 6, 0, 0, 3, 0, 1, 1,
            5, 0, 1, 2, 0, 0, 2, 1, 1, 1, 0)
n <- length(counts)  # Number of traps
sum_counts <- sum(counts)  # Total insects caught
```

# Posterior Grid Approximation

We approximate the posterior by evaluating it over a **grid of values** for \( (\theta_1, \theta_2) \):

```{r}
# Define grid for theta1 (log a) and theta2 (log b)
grid_theta1 <- seq(-2, 3, length.out = 100)
grid_theta2 <- seq(-2, 3, length.out = 100)
grid <- expand.grid(theta1 = grid_theta1, theta2 = grid_theta2)

grid$post_density <- apply(grid, 1, function(params) {
  theta1 <- params["theta1"]
  theta2 <- params["theta2"]
  a <- exp(theta1)
  b <- exp(theta2)
  sum(log(gamma(counts + a)) - log(gamma(a)) - lgamma(counts + 1) +
      a * log(b) - (counts + a) * log(b + 1)) - log(a * b)
})

grid$post_density <- grid$post_density - max(grid$post_density)
grid$post_density <- exp(grid$post_density)
grid$post_density <- grid$post_density / sum(grid$post_density)
```

# Simulating from the Posterior

We generate **1000 posterior samples** from the grid:

```{r}
set.seed(123)
posterior_samples <- grid[sample(1:nrow(grid), size = 1000, prob = grid$post_density, replace = TRUE), ]
```

# 90% Credible Intervals

We compute **90% credible intervals** for \(a\) and \(b\):

```{r}
ci_a <- quantile(exp(posterior_samples$theta1), c(0.05, 0.95))
ci_b <- quantile(exp(posterior_samples$theta2), c(0.05, 0.95))

list(a_CI = ci_a, b_CI = ci_b)
```

# Visualization of Posterior Distributions

```{r}
library(ggplot2)

data_posterior <- data.frame(a = exp(posterior_samples$theta1), b = exp(posterior_samples$theta2))

ggplot(data_posterior, aes(x = a)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(title = "Posterior Distribution of Parameter a",
       x = "a", y = "Density") +
  theme_minimal()

# Posterior distribution of b
ggplot(data_posterior, aes(x = b)) +
  geom_density(fill = "red", alpha = 0.4) +
  labs(title = "Posterior Distribution of Parameter b",
       x = "b", y = "Density") +
  theme_minimal()
```

# Summary of Results

- **90% Credible Interval for \(a\)**: `r round(ci_a, 3)`
- **90% Credible Interval for \(b\)**: `r round(ci_b, 3)`
- **Posterior Mean of \(a\)**: `r round(mean(exp(posterior_samples$theta1)), 3)`
- **Posterior Mean of \(b\)**: `r round(mean(exp(posterior_samples$theta2)), 3)`

