---
title: "Exercise 4.3: Bayesian Inference for Comparing Two Proportions"
author: "Rajitha M. Silva"
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this analysis, we compare the **risk of fatal accidents** between drivers **using seat belts** and those **not using seat belts**.

We model the number of fatal and nonfatal accidents using a **Bayesian approach** with **Beta priors**.

# Data

The observed accident data from the **Florida Highway Safety Department (1998)**:

```{r}
data_accidents <- data.frame(
  Safety_Equipment = c("None", "Seat Belt"),
  Fatal = c(1601, 510),
  Nonfatal = c(162527, 412368)
)

data_accidents$Total <- data_accidents$Fatal + data_accidents$Nonfatal
```

# Bayesian Model

We assume:

\[
Y_N \sim \text{Binomial}(n_N, p_N), \quad Y_S \sim \text{Binomial}(n_S, p_S)
\]

where:
- \( p_N \) = probability of fatality **without a seat belt**.
- \( p_S \) = probability of fatality **with a seat belt**.
- **Likelihood:** Binomial model for each group.
- **Prior:** Assume a **Beta(1,1) (Uniform) prior** for both probabilities.

# Posterior Distribution

Using Bayes' Theorem, the posterior distributions are:

\[
p_N | y_N \sim \text{Beta}(y_N + 1, n_N - y_N + 1)
\]

\[
p_S | y_S \sim \text{Beta}(y_S + 1, n_S - y_S + 1)
\]

# Posterior Simulation

We generate **10000 posterior samples** for \( p_N \) and \( p_S \):

```{r}
set.seed(123)

# Extract values
y_N <- data_accidents$Fatal[1]
n_N <- data_accidents$Total[1]
y_S <- data_accidents$Fatal[2]
n_S <- data_accidents$Total[2]

# Simulating posterior samples
n_samples <- 10000
p_N_samples <- rbeta(n_samples, y_N + 1, n_N - y_N + 1)
p_S_samples <- rbeta(n_samples, y_S + 1, n_S - y_S + 1)
```

# Relative Risk (RR) and Risk Difference (RD)

We compute the **relative risk (RR)** and **risk difference (RD)**:

```{r}
# Relative Risk (RR) = p_N / p_S
RR_samples <- p_N_samples / p_S_samples

# Risk Difference (RD) = p_N - p_S
RD_samples <- p_N_samples - p_S_samples
```

# Visualization of Posterior Distributions

```{r}
library(ggplot2)

data_posterior <- data.frame(
  Probability = c(p_N_samples, p_S_samples),
  Group = rep(c("No Seat Belt", "Seat Belt"), each = n_samples)
)

ggplot(data_posterior, aes(x = Probability, fill = Group)) +
  geom_density(alpha = 0.5) +
  labs(title = "Posterior Distributions of Fatality Probabilities",
       x = "Fatality Probability", y = "Density") +
  theme_minimal()
```

# 95% Credible Intervals

```{r}
ci_pN <- quantile(p_N_samples, c(0.025, 0.975))
ci_pS <- quantile(p_S_samples, c(0.025, 0.975))
ci_RR <- quantile(RR_samples, c(0.025, 0.975))
ci_RD <- quantile(RD_samples, c(0.025, 0.975))

list(
  pN_CI = ci_pN,
  pS_CI = ci_pS,
  RR_CI = ci_RR,
  RD_CI = ci_RD
)
```

# Posterior Probability that Risk is Higher Without Seat Belts

```{r}
prob_higher_risk <- mean(RD_samples > 0)
prob_higher_risk
```

# Summary of Results

- **95% Credible Interval for \( p_N \)** (No Seat Belt): `r round(ci_pN, 6)`
- **95% Credible Interval for \( p_S \)** (Seat Belt): `r round(ci_pS, 6)`
- **95% Credible Interval for Relative Risk (RR)**: `r round(ci_RR, 3)`
- **95% Credible Interval for Risk Difference (RD)**: `r round(ci_RD, 6)`
- **Probability that risk is higher without seat belts**: `r round(prob_higher_risk, 4)`
