---
title: "Known Probability Distributions"
author: "Rajitha M. Silva"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: false
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Learning objectives

By the end of this lesson, students should be able to:

1. distinguish between discrete and continuous random variables;
2. identify whether a distribution is described by a pmf or a pdf;
3. recognize common probability distributions and their parameters;
4. compute probabilities using R;
5. interpret the mean and variance of each distribution.

---

## Random variables and probability distributions

A **random variable** is a numerical quantity whose value is determined by the outcome of a random experiment. We usually write random variables using capital letters such as \(X\), \(Y\), and \(Z\), and possible observed values using lower-case letters such as \(x\), \(y\), and \(z\).

Examples:

- Toss a coin once and define \(X=1\) if a head appears and \(X=0\) otherwise.
- Count the number of cars passing a toll booth in one hour.
- Measure the waiting time until the next customer arrives.

A **probability distribution** describes how probability is assigned to the possible values of a random variable.

---

## Discrete and continuous random variables

### Discrete random variables

A random variable is **discrete** if it takes countable values, usually integers.

Examples:

- number of correct answers in an MCQ test;
- number of accidents in a month;
- number of calls received in an hour.

For a discrete random variable, probabilities are described by a **probability mass function** (pmf)

\[
p(x)=P(X=x).
\]

A valid pmf must satisfy

\[
p(x)\ge 0, \qquad \sum_x p(x)=1.
\]

### Continuous random variables

A random variable is **continuous** if it can take any value in an interval of the real line.

Examples:

- height;
- weight;
- lifetime of a machine;
- waiting time.

For a continuous random variable, probabilities are described by a **probability density function** (pdf) \(f(x)\). A valid pdf satisfies

\[
f(x)\ge 0, \qquad \int_{-\infty}^{\infty} f(x)\,dx=1.
\]

For continuous random variables,

\[
P(a<X<b)=\int_a^b f(x)\,dx,
\]

and \(P(X=x)=0\) for any single exact value \(x\).

---

## Summary of common distributions

| Distribution | Type | Support | Parameter(s) | Mean | Variance | R functions |
|---|---|---:|---|---:|---:|---|
| Bernoulli | Discrete | \(0,1\) | \(p\) | \(p\) | \(p(1-p)\) | `dbinom`, `pbinom`, `rbinom` |
| Binomial | Discrete | \(0,1,\ldots,n\) | \(n,p\) | \(np\) | \(np(1-p)\) | `dbinom`, `pbinom`, `rbinom` |
| Poisson | Discrete | \(0,1,2,\ldots\) | \(\lambda\) | \(\lambda\) | \(\lambda\) | `dpois`, `ppois`, `rpois` |
| Uniform | Continuous | \((a,b)\) | \(a,b\) | \((a+b)/2\) | \((b-a)^2/12\) | `dunif`, `punif`, `runif` |
| Exponential | Continuous | \((0,\infty)\) | \(\lambda\) | \(1/\lambda\) | \(1/\lambda^2\) | `dexp`, `pexp`, `rexp` |
| Normal | Continuous | \((-\infty,\infty)\) | \(\mu,\sigma^2\) | \(\mu\) | \(\sigma^2\) | `dnorm`, `pnorm`, `rnorm` |

In R, distribution functions commonly use prefixes:

- `d`: density or mass function;
- `p`: cumulative distribution function;
- `q`: quantile function;
- `r`: random generation.

---

# Discrete probability distributions

## Bernoulli distribution

A random variable \(X\) follows a **Bernoulli distribution** with parameter \(p\), written

\[
X \sim \mathrm{Bernoulli}(p),
\]

if

\[
P(X=1)=p, \qquad P(X=0)=1-p, \qquad 0<p<1.
\]

The pmf can be written compactly as

\[
P(X=x)=p^x(1-p)^{1-x}, \qquad x=0,1.
\]

Mean and variance:

\[
E(X)=p, \qquad \mathrm{Var}(X)=p(1-p).
\]

### Example: one coin toss

Let \(X=1\) if a fair coin shows head and \(X=0\) otherwise. Then \(X\sim \mathrm{Bernoulli}(0.5)\).

```{r bernoulli-plot}
p <- 0.5
bernoulli_df <- data.frame(
  x = c(0, 1),
  prob = dbinom(c(0, 1), size = 1, prob = p)
)

ggplot(bernoulli_df, aes(x = factor(x), y = prob)) +
  geom_col(width = 0.5) +
  labs(title = "Bernoulli(0.5) distribution",
       x = "x", y = "P(X = x)") +
  theme_minimal()
```

---

## Binomial distribution

A random variable \(X\) follows a **Binomial distribution**, written

\[
X\sim \mathrm{Binomial}(n,p),
\]

if \(X\) counts the number of successes in \(n\) independent Bernoulli trials, each with success probability \(p\).

The pmf is

\[
P(X=x)=\binom{n}{x}p^x(1-p)^{n-x}, \qquad x=0,1,\ldots,n.
\]

Mean and variance:

\[
E(X)=np, \qquad \mathrm{Var}(X)=np(1-p).
\]

### Example: random guessing in MCQs

Suppose there are 3 multiple-choice questions, each with 4 options and only one correct answer. A student guesses each answer independently. Let \(X\) be the number of correct answers. Then

\[
X\sim \mathrm{Binomial}(3,0.25).
\]

```{r binomial-example}
n <- 3
p <- 0.25
x <- 0:n
binom_df <- data.frame(
  x = x,
  prob = dbinom(x, size = n, prob = p)
)

binom_df
```

```{r binomial-plot}
ggplot(binom_df, aes(x = x, y = prob)) +
  geom_col(width = 0.5) +
  scale_x_continuous(breaks = x) +
  labs(title = "Binomial distribution: n = 3, p = 0.25",
       x = "Number of correct answers", y = "P(X = x)") +
  theme_minimal()
```

The probability of exactly two correct answers is

```{r binomial-probability}
dbinom(2, size = 3, prob = 0.25)
```

The probability of at least one correct answer is

```{r binomial-at-least-one}
1 - dbinom(0, size = 3, prob = 0.25)
# equivalently: 1 - pbinom(0, size = 3, prob = 0.25)
```

---

## Poisson distribution

A random variable \(X\) follows a **Poisson distribution**, written

\[
X\sim \mathrm{Poisson}(\lambda),
\]

if

\[
P(X=x)=\frac{e^{-\lambda}\lambda^x}{x!}, \qquad x=0,1,2,\ldots, \quad \lambda>0.
\]

Mean and variance:

\[
E(X)=\lambda, \qquad \mathrm{Var}(X)=\lambda.
\]

The Poisson distribution is commonly used to model counts of events occurring in a fixed interval of time or space, such as calls per hour, defects per item, or errors per page.

### Example: errors on a page

Suppose the average number of errors per page is \(0.2\). Let \(X\) be the number of errors on a randomly selected page. Then

\[
X\sim \mathrm{Poisson}(0.2).
\]

Probability of no error:

```{r poisson-zero}
dpois(0, lambda = 0.2)
```

Probability of two or more errors:

```{r poisson-two-or-more}
1 - ppois(1, lambda = 0.2)
```

```{r poisson-plot}
lambda <- 0.2
pois_df <- data.frame(
  x = 0:8,
  prob = dpois(0:8, lambda = lambda)
)

ggplot(pois_df, aes(x = x, y = prob)) +
  geom_col(width = 0.6) +
  scale_x_continuous(breaks = 0:8) +
  labs(title = expression(Poisson(lambda == 0.2)),
       x = "Number of errors", y = "P(X = x)") +
  theme_minimal()
```

---

# Continuous probability distributions

## Uniform distribution

A random variable \(X\) follows a **Uniform distribution** on \((a,b)\), written

\[
X\sim U(a,b),
\]

if its pdf is

\[
f(x)=\frac{1}{b-a}, \qquad a<x<b.
\]

Mean and variance:

\[
E(X)=\frac{a+b}{2}, \qquad \mathrm{Var}(X)=\frac{(b-a)^2}{12}.
\]

### Example

Let \(X\sim U(0,10)\). Find \(P(2<X<6)\).

\[
P(2<X<6)=\frac{6-2}{10-0}=0.4.
\]

```{r uniform-example}
punif(6, min = 0, max = 10) - punif(2, min = 0, max = 10)
```

```{r uniform-plot}
unif_df <- data.frame(x = seq(-1, 11, length.out = 400)) |>
  mutate(density = dunif(x, min = 0, max = 10))

ggplot(unif_df, aes(x = x, y = density)) +
  geom_line(linewidth = 1) +
  labs(title = "Uniform distribution on (0, 10)",
       x = "x", y = "f(x)") +
  theme_minimal()
```

---

## Exponential distribution

A random variable \(X\) follows an **Exponential distribution**, written

\[
X\sim \mathrm{Exponential}(\lambda),
\]

if its pdf is

\[
f(x)=\lambda e^{-\lambda x}, \qquad x>0, \quad \lambda>0.
\]

Mean and variance:

\[
E(X)=\frac{1}{\lambda}, \qquad \mathrm{Var}(X)=\frac{1}{\lambda^2}.
\]

The exponential distribution is often used to model waiting time until the next event.

Its cdf is

\[
F(x)=P(X\le x)=1-e^{-\lambda x}, \qquad x>0.
\]

Therefore,

\[
P(X>x)=e^{-\lambda x}.
\]

### Example: machine repair time

Suppose the average repair time of a machine is 0.5 hours. Since \(E(X)=1/\lambda\), we have

\[
\lambda=\frac{1}{0.5}=2.
\]

Find \(P(X>2)\).

\[
P(X>2)=e^{-2(2)}=e^{-4}.
\]

```{r exponential-example}
1 - pexp(2, rate = 2)
```

```{r exponential-plot}
exp_df <- data.frame(x = seq(0, 4, length.out = 400)) |>
  mutate(density = dexp(x, rate = 2))

ggplot(exp_df, aes(x = x, y = density)) +
  geom_line(linewidth = 1) +
  labs(title = expression(Exponential(lambda == 2)),
       x = "Repair time", y = "f(x)") +
  theme_minimal()
```

---

## Normal distribution

A random variable \(X\) follows a **Normal distribution**, written

\[
X\sim N(\mu,\sigma^2),
\]

if its pdf is

\[
f(x)=\frac{1}{\sigma\sqrt{2\pi}}\exp\left[-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2\right],
\qquad -\infty<x<\infty.
\]

Mean and variance:

\[
E(X)=\mu, \qquad \mathrm{Var}(X)=\sigma^2.
\]

Important characteristics:

- The curve is bell-shaped and symmetric about \(\mu\).
- Mean = median = mode.
- Approximately 68.27% of observations lie within \(\mu\pm \sigma\).
- Approximately 95.45% lie within \(\mu\pm 2\sigma\).
- Approximately 99.73% lie within \(\mu\pm 3\sigma\).

### Standard normal transformation

If

\[
X\sim N(\mu,\sigma^2),
\]

then

\[
Z=\frac{X-\mu}{\sigma}\sim N(0,1).
\]

This transformation allows probabilities for any normal random variable to be computed using the standard normal distribution.

### Example: weekly TV viewers

Suppose the number of viewers of a TV show per week, in millions, follows

\[
X\sim N(29,5^2).
\]

Find the probability that next week's viewership is between 30 and 34 million.

```{r normal-between}
pnorm(34, mean = 29, sd = 5) - pnorm(30, mean = 29, sd = 5)
```

Find the probability that next week's viewership is at least 23 million.

```{r normal-at-least}
1 - pnorm(23, mean = 29, sd = 5)
```

Find the probability that next week's viewership exceeds 40 million.

```{r normal-exceed}
1 - pnorm(40, mean = 29, sd = 5)
```

```{r normal-plot}
norm_df <- data.frame(x = seq(10, 50, length.out = 600)) |>
  mutate(density = dnorm(x, mean = 29, sd = 5))

ggplot(norm_df, aes(x = x, y = density)) +
  geom_line(linewidth = 1) +
  labs(title = expression(Normal(mu == 29, sigma == 5)),
       x = "Viewers per week (millions)", y = "f(x)") +
  theme_minimal()
```

---

## Comparing distributions by simulation

Simulation helps students connect formulas with long-run frequency behavior.

```{r simulation-comparison}
set.seed(502)

sim_df <- tibble(
  Bernoulli = rbinom(1000, size = 1, prob = 0.5),
  Binomial = rbinom(1000, size = 10, prob = 0.4),
  Poisson = rpois(1000, lambda = 3),
  Uniform = runif(1000, min = 0, max = 1),
  Exponential = rexp(1000, rate = 2),
  Normal = rnorm(1000, mean = 0, sd = 1)
)

summary(sim_df)
```

```{r simulation-normal}
ggplot(sim_df, aes(x = Normal)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_density(linewidth = 1) +
  labs(title = "Simulated standard normal observations",
       x = "x", y = "Density") +
  theme_minimal()
```

---

## Classroom discussion questions

1. What is the difference between a probability mass function and a probability density function?
2. Why is \(P(X=x)=0\) for a continuous random variable?
3. When is the Binomial distribution appropriate?
4. Why does the Poisson distribution have the same mean and variance?
5. What does the parameter \(\lambda\) represent in the Exponential distribution?
6. Why do we standardize a normal random variable?

---

## Practice problems

### Problem 1: Binomial

A student answers 8 true/false questions by guessing. Let \(X\) be the number of correct answers.

1. Identify the distribution of \(X\).
2. Find \(P(X=5)\).
3. Find \(P(X\ge 6)\).

```{r practice-binomial, eval=FALSE}
dbinom(5, size = 8, prob = 0.5)
1 - pbinom(5, size = 8, prob = 0.5)
```

### Problem 2: Poisson

A call center receives an average of 4 calls per minute. Let \(X\) be the number of calls in one minute.

1. Identify the distribution of \(X\).
2. Find \(P(X=3)\).
3. Find \(P(X>5)\).

```{r practice-poisson, eval=FALSE}
dpois(3, lambda = 4)
1 - ppois(5, lambda = 4)
```

### Problem 3: Exponential

The waiting time until the next bus follows an exponential distribution with mean 10 minutes.

1. Find \(\lambda\).
2. Find the probability that the waiting time exceeds 15 minutes.

```{r practice-exponential, eval=FALSE}
rate <- 1 / 10
1 - pexp(15, rate = rate)
```

### Problem 4: Normal

Suppose exam scores follow \(N(70,10^2)\).

1. Find \(P(X>85)\).
2. Find \(P(60<X<80)\).
3. Find the 90th percentile.

```{r practice-normal, eval=FALSE}
1 - pnorm(85, mean = 70, sd = 10)
pnorm(80, mean = 70, sd = 10) - pnorm(60, mean = 70, sd = 10)
qnorm(0.90, mean = 70, sd = 10)
```

---


