---
title: "Exercise 4.2: Bayesian Inference for the Behrens-Fisher Problem"
author: "Rajitha M. Silva"
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The **Behrens-Fisher problem** involves comparing the means of two independent normal distributions when their variances are unknown. 
This is relevant in various scientific applications, such as comparing measurements from different populations.

Here, we analyze the **mandible lengths of male and female golden jackals** to determine whether the means differ significantly.

# Data

The observed mandible lengths (in mm) are:

```{r}
male_lengths <- c(120, 107, 110, 116, 114, 111, 113, 117, 114, 112)
female_lengths <- c(110, 111, 107, 108, 110, 105, 107, 106, 111, 111)
```

# Bayesian Model

We assume:

\[
X_i \sim N(\mu_1, \sigma_1^2), \quad Y_j \sim N(\mu_2, \sigma_2^2)
\]

for two independent normal populations.

## Prior Distribution

We use the **noninformative prior**:

\[
p(\mu_1, \sigma_1^2, \mu_2, \sigma_2^2) \propto \frac{1}{\sigma_1^2 \sigma_2^2}
\]

which expresses **no strong prior knowledge** about the parameters.

## Posterior Distribution

From Bayesian theory, we derive the posterior distributions:

1. **Conditional posterior** of \( \mu_1 \) given \( \sigma_1^2 \):
   
   \[
   \mu_1 | \sigma_1^2, X \sim N(\bar{X}, \frac{\sigma_1}{\sqrt{m}})
   \]

2. **Marginal posterior** of \( \sigma_1^2 \):
   
   \[
   \sigma_1^2 | X \sim S_1 \cdot \chi^{-2}_{m-1}
   \]

where \( S_1 \) is the sum of squared deviations from the mean.

The same holds for \( \mu_2 \) and \( \sigma_2^2 \) in the second population.

# Posterior Simulation

To generate samples from the **joint posterior** of \( (\mu_1, \sigma_1^2, \mu_2, \sigma_2^2) \), we:
1. Compute summary statistics.
2. Sample \( \sigma_1^2 \) and \( \sigma_2^2 \) from inverse chi-square distributions.
3. Sample \( \mu_1 \) and \( \mu_2 \) from normal distributions.

```{r}
set.seed(123)

# Compute summary statistics
m <- length(male_lengths)
n <- length(female_lengths)

x_bar <- mean(male_lengths)
y_bar <- mean(female_lengths)

S1 <- sum((male_lengths - x_bar)^2)
S2 <- sum((female_lengths - y_bar)^2)

# Simulate from posterior distribution
sigma1_sq_samples <- S1 / rchisq(1000, df = m - 1)
sigma2_sq_samples <- S2 / rchisq(1000, df = n - 1)

mu1_samples <- rnorm(1000, mean = x_bar, sd = sqrt(sigma1_sq_samples) / sqrt(m))
mu2_samples <- rnorm(1000, mean = y_bar, sd = sqrt(sigma2_sq_samples) / sqrt(n))
```

# Difference in Means

We estimate the difference in population means:

```{r}
diff_mu_samples <- mu1_samples - mu2_samples

# Compute posterior mean and 95% credible interval
posterior_mean_diff <- mean(diff_mu_samples)
ci_diff <- quantile(diff_mu_samples, c(0.025, 0.975))

list(mean_diff = posterior_mean_diff, credible_interval = ci_diff)
```

# 90% Credible Intervals

## 90% Credible Interval for \( \mu_1 \) and \( \mu_2 \)

```{r}
ci_mu1 <- quantile(mu1_samples, c(0.05, 0.95))
ci_mu2 <- quantile(mu2_samples, c(0.05, 0.95))

list(mu1_ci = ci_mu1, mu2_ci = ci_mu2)
```

## 90% Credible Interval for \( \sigma_1 \) and \( \sigma_2 \)

```{r}
sigma1_samples <- sqrt(sigma1_sq_samples)
sigma2_samples <- sqrt(sigma2_sq_samples)

ci_sigma1 <- quantile(sigma1_samples, c(0.05, 0.95))
ci_sigma2 <- quantile(sigma2_samples, c(0.05, 0.95))

list(sigma1_ci = ci_sigma1, sigma2_ci = ci_sigma2)
```

# Summary of Results

- **95% Credible Interval for Difference in Means**: `r round(ci_diff, 2)`
- **90% Credible Interval for \( \mu_1 \)**: `r round(ci_mu1, 2)`
- **90% Credible Interval for \( \mu_2 \)**: `r round(ci_mu2, 2)`
- **90% Credible Interval for \( \sigma_1 \)**: `r round(ci_sigma1, 2)`
- **90% Credible Interval for \( \sigma_2 \)**: `r round(ci_sigma2, 2)`
