
---
title: "Week8: Essential Non-parametric One-Sample Procedures in R"

output: 
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1  Motivation

Classical t-procedures rely on Normality; when that assumption is shaky—or when data are merely ordinal—**rank-based methods** step in.  
This note reviews eight core ideas:

1. One-sample inference on a proportion  
2. The sign test  
3. Wilcoxon signed-rank test  
4. Hodges–Lehmann estimator & CIs  
5. Adjustments for ties  
6. Asymptotic large-sample versions  
7. Bootstrap confidence intervals & tests  
8. A glimpse at robustness

---

## 2  One-Sample Proportion Problems

Let \\(X \\sim \\text{Bernoulli}(p)\\). For a sample \\(X_1,\\dots,X_n\\), the number of successes \\(S = \\sum X_i\\) follows \\(\\text{Bin}(n, p)\\).

#### 2.1  Point & Interval Estimation  

\[
\hat{p} = \frac{S}{n}, \quad 
\text{CI}_{\text{Wald}} = \hat{p} \pm z_{1 - \alpha/2} \sqrt{ \frac{ \hat{p}(1 - \hat{p}) }{n} }
\]


```{r prop-ci}
phat <- 10 / 143
se <- sqrt(phat * (1 - phat) / 143)
phat + c(-1, 1) * qnorm(0.975) * se
```

---

## 3  The Sign Test

The **sign test** is a simple non-parametric method to test whether the median of a population is equal to a specified value (often 0). It works by counting how many observations are above and below the hypothesized median.

The **test statistic** is defined as:

```{=latex}
\[
S^+ = \#\{X_i > 0\}
\]
```

### Explanation:

- \( S^+ \): This is the **number of positive signs** in the sample — i.e., how many observations are greater than the hypothesized median (usually 0).
- \( \#\{X_i > 0\} \): The symbol “\#” denotes the count of elements satisfying the condition inside the set brackets. So this represents “the number of \( X_i \) such that \( X_i > 0 \)”.

In simple terms:  
> Count how many observations are positive after subtracting the hypothesized median. That count is \( S^+ \).

The test assumes that under the null hypothesis (e.g., median = 0), positive and negative signs should be equally likely.

```{r sign-example}
# Example data: differences from some hypothesized value (e.g., 0)
x <- c(3.2, -1.1, 4.7, 0, -2.5, 5.0)

# Count of values greater than 0 (ignores zeros)
S_plus <- sum(x > 0)
S_plus
```

If the number of positives is significantly more (or less) than expected under randomness, we reject the null hypothesis.


```{r sign}
x <- c(3.2, -1.1, 4.7, 0, -2.5, 5.0)
Splus <- sum(x > 0)
pval <- 1 - pbinom(Splus - 1, length(x), 0.5)
pval
```

---

## Wilcoxon Signed-Rank Test

### Test Statistic

1. Remove zeros, take absolute values, rank them  
2. \\( W^+ = \\sum_{X_i > 0} \\text{rank}(|X_i|) \\)  

### 4.2 Example – Nursery-school Twins

```{r twins}
school <- c(82, 69, 73, 43, 58, 56, 76, 65)
home <- c(63, 42, 74, 37, 51, 43, 80, 62)
resp <- school - home
wilcox.test(resp, alternative = "greater", conf.int = TRUE)
```


## 5 Confidence Intervals & Standard Errors

### 5.1 Distribution-free CI for \\( \\theta \\) using Walsh Averages

Use \\( A_{(1)} \\le A_{(2)} \\le \\dots \\le A_{(m)} \\), where \\( m = n(n+1)/2 \\).  
A 95% CI is \\( (A_{(c+1)}, A_{(m-c)}) \\) for appropriate critical value \\( c \\).

### 5.2 Standard Error via CI Length

\\[
\\widehat{\\text{SE}}(\\hat\\theta_{\\mathrm{HL}}) = \\frac{A_{(m - c)} - A_{(c+1)}}{2 z_{1 - \\alpha/2} \\sqrt{n}}
\\]

---

## 6  Handling Ties

- Zeros are omitted in both sign and Wilcoxon tests  
- If ties exist, assign average ranks  
- Adjust Wilcoxon variance:

\\[
\\operatorname{Var}(W^+) = \\frac{n(n+1)(2n+1)}{24} - \\frac{1}{2} \\sum t_j(t_j - 1)(t_j + 1)
\\]

where \\( t_j \\) is the size of tie group \\( j \\)

---

## 7  Large Sample Approximations

For \\( n \\geq 20 \\), use Normal approximations:

\\[
Z_W = \\frac{W^+ - n(n+1)/4}{\\sqrt{\\operatorname{Var}(W^+)}},\\quad
Z_S = \\frac{S^+ - n/2}{\\sqrt{n/4}}
\\]

Include continuity correction ±0.5 when appropriate.

---

## 8  Bootstrap Confidence Intervals & Tests

Percentile CI approach:

```{r bootstrap-ci}
set.seed(123)
x <- rnorm(25, 30, 5)
B <- 2000
theta_star <- replicate(B, mean(sample(x, replace = TRUE)))
quantile(theta_star, c(0.025, 0.975))
```

Bootstrap tests resample under \\( H_0 \\): center data at \\( \\theta_0 \\), recompute test statistic, and compare to observed.

---


