---
title: "Practical: F-Statistic and ANOVA Exercises"
author: "Dr. Rajitha M. Silva"
output: html_document

---

# Objectives

1. Apply One-Way ANOVA to simulated sports data.  
2. Interpret ANOVA tables and F-Statistic results.  
3. Perform residual diagnostics for ANOVA models.  
4. Conduct Two-Way ANOVA with interaction and interpret main and interaction effects.  
5. Show equivalent Python code (with `eval=FALSE`) for each task.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(2025)
```

# 1. One-Way ANOVA Practical

## 1.1 Simulate Batting Average Data

Simulate batting averages for three teams (A, B, C), 10 players each.

```{r simulate-batting}
team_A <- rnorm(10, mean = 0.280, sd = 0.020)
team_B <- rnorm(10, mean = 0.270, sd = 0.018)
team_C <- rnorm(10, mean = 0.290, sd = 0.022)

batting_df <- data.frame(
  average = c(team_A, team_B, team_C),
  team = factor(rep(c("A", "B", "C"), each = 10))
)

# Preview data
head(batting_df)
```

**Task 1**: Compute group means and the grand mean manually. Show your calculations in R.

```{r group-means}
# Compute group and grand means
group_means <- tapply(batting_df$average, batting_df$team, mean)
grand_mean <- mean(batting_df$average)
group_means
grand_mean
```

## 1.2 Calculate SSB, SSW, MSB, MSW, and F-Statistic Manually

Use formulas:

- SSB = sum_{j} n_j * (mean_j - grand_mean)^2  
- SSW = sum_{j,i} (Y_ij - mean_j)^2  

```{r manual-ssb-ssw}
n_j <- table(batting_df$team)
ssb <- sum(n_j * (group_means - grand_mean)^2)
ssw <- sum(tapply((batting_df$average - rep(group_means, each = 10))^2,
                  batting_df$team, sum))
msb <- ssb / (length(group_means) - 1)
msw <- ssw / (nrow(batting_df) - length(group_means))
F_value <- msb / msw

ssb
ssw
msb
msw
F_value
```

**Task 2**: Compare your manual F-Statistic to the value from `aov()`.

```{r anova-compare}
anova_result <- aov(average ~ team, data = batting_df)
anova_summary <- summary(anova_result)
anova_summary
```

## 1.3 Residual Diagnostics

Perform residual diagnostics on the One-Way ANOVA model.

```{r diagnostics}
# Extract residuals and fitted values
resid_vals <- residuals(anova_result)
fitted_vals <- fitted(anova_result)

# Q-Q plot
qqnorm(resid_vals); qqline(resid_vals, col = "blue", lwd = 2)

# Residuals vs Fitted
plot(fitted_vals, resid_vals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lwd = 2)
```

**Task 3**: Evaluate whether ANOVA assumptions (normality, homoscedasticity) hold based on the plots.

## 1.4 Post-Hoc Test: Tukey HSD

Apply Tukey HSD to the ANOVA model.

```{r tukey-practical}
tukey_results <- TukeyHSD(anova_result)
tukey_results
plot(tukey_results)
```

**Task 4**: Identify which pairs of teams have significantly different batting averages at α = 0.05.

## 1.5 Equivalent Python Code (No Execution)

```{python python-oneway, eval=FALSE}
import pandas as pd
import numpy as np
import statsmodels.api as sm
from statsmodels.formula.api import ols
from statsmodels.stats.multicomp import pairwise_tukeyhsd

# Create DataFrame
np.random.seed(2025)
team_A_py = np.random.normal(loc=0.280, scale=0.020, size=10)
team_B_py = np.random.normal(loc=0.270, scale=0.018, size=10)
team_C_py = np.random.normal(loc=0.290, scale=0.022, size=10)

batting_py = pd.DataFrame({
    "average": np.concatenate([team_A_py, team_B_py, team_C_py]),
    "team": ["A"]*10 + ["B"]*10 + ["C"]*10
})

# Fit One-Way ANOVA
model = ols('average ~ C(team)', data=batting_py).fit()
anova_table = sm.stats.anova_lm(model, typ=1)
print(anova_table)

# Residual diagnostics and Tukey HSD
tukey = pairwise_tukeyhsd(endog=batting_py["average"],
                          groups=batting_py["team"],
                          alpha=0.05)
print(tukey)
```

# 2. Two-Way ANOVA Practical

## 2.1 Simulate Batting Data by Team and Order

Simulate batting averages for teams (A, B, C) and batting orders (Top, Middle, Bottom), with 10 observations per combination (balanced design).

```{r simulate-two-way}
teams <- rep(c("A", "B", "C"), each = 30)
orders <- rep(rep(c("Top", "Middle", "Bottom"), each = 10), times = 3)

team_effects <- c(A = 0.000, B = -0.005, C = 0.005)
order_effects <- c(Top = 0.010, Middle = 0.000, Bottom = -0.010)

batting_two_df <- data.frame(
  average = 0.280 +
    team_effects[teams] +
    order_effects[orders] +
    rnorm(90, mean = 0, sd = 0.015),
  team = factor(teams),
  order = factor(orders)
)

# Preview data
head(batting_two_df)
table(batting_two_df$team, batting_two_df$order)
```

## 2.2 Fit Two-Way ANOVA Model

Fit a Two-Way ANOVA with interaction: `average ~ team * order`.

```{r anova-two-way-practical}
anova_two <- aov(average ~ team * order, data = batting_two_df)
summary(anova_two)
```

**Task 5**: Report which main effects (team, order) and the interaction are significant at α = 0.05.

## 2.3 Interaction Plot

Produce an interaction plot to visualize team by order effects.

```{r interaction-plot}
interaction.plot(x.factor = batting_two_df$order,
                 trace.factor = batting_two_df$team,
                 response = batting_two_df$average,
                 fun = mean,
                 xlab = "Batting Order",
                 ylab = "Mean Batting Average",
                 trace.label = "Team")
```

**Task 6**: Interpret the interaction plot. Does the effect of batting order on average depend on the team?

## 2.4 Equivalent Python Code (No Execution)

```{python python-two-way-practical, eval=FALSE}
import pandas as pd
import numpy as np
import statsmodels.api as sm
from statsmodels.formula.api import ols

# Create DataFrame
np.random.seed(2025)
teams_py = np.repeat(["A", "B", "C"], 30)
orders_py = np.tile(np.repeat(["Top", "Middle", "Bottom"], 10), 3)

team_effects_py = {"A": 0.000, "B": -0.005, "C": 0.005}
order_effects_py = {"Top": 0.010, "Middle": 0.000, "Bottom": -0.010}

batting_two_py = pd.DataFrame({
    "average": 0.280 +
        [team_effects_py[t] for t in teams_py] +
        [order_effects_py[o] for o in orders_py] +
        np.random.normal(loc=0, scale=0.015, size=90),
    "team": teams_py,
    "order": orders_py
})

# Fit Two-Way ANOVA
model_two = ols('average ~ C(team) * C(order)', data=batting_two_py).fit()
anova_two_table = sm.stats.anova_lm(model_two, typ=2)
print(anova_two_table)
```

# 3. Submission Instructions

- Knit this R Markdown to HTML and submit both the HTML and the `.Rmd` file.  
- Include answers to Tasks 1–6 in the final document.  
- For Python code chunks, you may copy them into a `.py` file if needed, but no execution is required.
