
---
title: "Chi-Square Test - Practical Session"
author: "Dr. Rajitha M. Silva"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Practical Objective

This session helps students understand the computation and interpretation of the chi-square test, including both theoretical and resampling-based approaches, with applications to sports analytics.

# Task 1: Create the Contingency Table

**Scenario:**  
Three volleyball players attempted 50 serves each. The number of successful and failed serves is recorded:

| Player   | Success | Fail |
|----------|---------|------|
| Player A | 42      | 8    |
| Player B | 36      | 14   |
| Player C | 40      | 10   |

## 1.1 R: Create the matrix

```{r}
observed <- matrix(
  c(42, 8,
    36, 14,
    40, 10),
  nrow = 3,
  byrow = TRUE
)
rownames(observed) <- c("Player A", "Player B", "Player C")
colnames(observed) <- c("Success", "Fail")
observed
```

## 1.2 Python: Create the table (display only)

```{python, eval=FALSE}
import numpy as np
observed = np.array([[42, 8],
                     [36, 14],
                     [40, 10]])
print(observed)
```

# Task 2: Compute Expected Counts and Chi-Square Statistic

## 2.1 R: Manually compute expected values and Pearson residuals

```{r}
row_totals <- rowSums(observed)
col_totals <- colSums(observed)
grand_total <- sum(observed)

expected <- outer(row_totals, col_totals) / grand_total
pearson_residuals <- (observed - expected) / sqrt(expected)
chi2_manual <- sum(pearson_residuals^2)

expected
pearson_residuals
chi2_manual
```

## 2.2 Python: Display equivalent code (not run)

```{python, eval=FALSE}
row_totals = observed.sum(axis=1)
col_totals = observed.sum(axis=0)
grand_total = observed.sum()

expected = np.outer(row_totals, col_totals) / grand_total
residuals = (observed - expected) / np.sqrt(expected)
chi2_stat = np.sum(residuals**2)
```

# Task 3: Theoretical Chi-Square Test

## 3.1 R

```{r}
chisq.test(observed)
```

## 3.2 Python (code only)

```{python, eval=FALSE}
from scipy.stats import chi2_contingency
chi2, p, df, expected = chi2_contingency(observed)
```

# Task 4: Permutation Test (Resampling Approach)

## 4.1 R

```{r}
set.seed(1234)
box <- c(rep(1, sum(observed[, "Success"])), rep(0, sum(observed[, "Fail"])))
perm_chi2 <- function(box, row_sizes, col_totals) {
  shuffled <- sample(box)
  idx1 <- shuffled[1:row_sizes[1]]
  idx2 <- shuffled[(row_sizes[1]+1):(sum(row_sizes[1:2]))]
  idx3 <- shuffled[(sum(row_sizes[1:2])+1):sum(row_sizes)]
  made_counts <- c(sum(idx1), sum(idx2), sum(idx3))
  missed_counts <- row_sizes - made_counts
  perm_obs <- cbind(made_counts, missed_counts)
  perm_expected <- outer(row_sizes, col_totals) / sum(row_sizes)
  resids <- (perm_obs - perm_expected) / sqrt(perm_expected)
  sum(resids^2)
}

row_sizes <- row_totals
perm_stats <- replicate(2000, perm_chi2(box, row_sizes, col_totals))
mean(perm_stats >= chi2_manual)
```

# Task 5: Interpretation Questions

1. Which cells contribute most to the chi-square statistic?
2. Does the permutation p-value differ significantly from the theoretical p-value?
3. Are you confident rejecting the null hypothesis? Why or why not?

# Task 6: Optional Fisher’s Exact Test for 2x2

## R Only:

```{r}
fisher.test(observed[1:2, ])
```

---

**Submission Instructions:**  
- Knit your `.Rmd` file to HTML and submit it on the LMS.
- Python users may submit a Jupyter Notebook version separately.
