---
title: "Exercise 4.8: Bayesian Logistic Regression for ACT Scores and Success Prediction"
author: "Rajitha M. Silva"
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this exercise, we analyze the relationship between students' **ACT scores** and their success (achieving an A or B) in a **business calculus class** using **Bayesian logistic regression**. The success probability follows the **logistic model**:

\[
\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 x_i
\]

where:
- \( p_i \) is the probability of success for a student with **ACT score** \( x_i \),
- \( \beta_0 \) and \( \beta_1 \) are **regression parameters** to be estimated.

# Problem Setup

The math department collected data for **students' ACT scores** and their **success rates**:

```{r}
data <- data.frame(
  ACT_Score = c(16, 18, 20, 22, 24, 26, 28),
  Students = c(2, 7, 14, 26, 13, 14, 3),
  Successes = c(0, 0, 6, 12, 7, 9, 3)
)
```

# Prior Specification using Conditional Means Prior

The department has prior beliefs about the probability of success for certain ACT scores:

- **ACT = 18:** Quartiles for probability = (0.15, 0.35)
- **ACT = 26:** Quartiles for probability = (0.75, 0.95)

Using **beta.select**, we match these beliefs to Beta distributions:

```{r}
library(LearnBayes)

# Compute prior parameters using beta.select
prior_18 <- beta.select(list(p = 0.25, x = 0.15), list(p = 0.75, x = 0.35))
prior_26 <- beta.select(list(p = 0.25, x = 0.75), list(p = 0.75, x = 0.95))

prior_18  # Beta parameters for ACT = 18
prior_26  # Beta parameters for ACT = 26
```

# Posterior Distribution of \( \beta_0, \beta_1 \)

We define the likelihood using the **logisticpost** function and visualize the **posterior region**:

```{r}
# Define the data matrix for likelihood calculation
data_matrix <- cbind(data$ACT_Score, data$Students, data$Successes)

# Define prior data based on the matched Beta parameters
prior_data <- rbind(
  c(18, sum(prior_18), prior_18[1]),
  c(26, sum(prior_26), prior_26[1])
)

data_combined <- rbind(data_matrix, prior_data)
```

Now, we compute the **posterior distribution** of \( \beta_0, \beta_1 \) using **contour plots**:

```{r}
# Contour plot of posterior distribution
mycontour(logisticpost, c(-10, 5, -5, 3), data_combined, xlab="beta0", ylab="beta1")
```

# Simulating from the Posterior Distribution

We now generate **1000 posterior samples**:

```{r}
set.seed(123)
posterior_samples <- simcontour(logisticpost, data_combined, 1000)
```

We overlay the **posterior samples** on the contour plot:

```{r}
points(posterior_samples)
```

# Predicting Success Probability for ACT = 20

To estimate the **probability of success** for a student with an **ACT score of 20**, we compute:

\[
p_{20} = \frac{\exp(\beta_0 + 20\beta_1)}{1 + \exp(\beta_0 + 20\beta_1)}
\]

```{r}
p_success_20 <- exp(posterior_samples$x + 20 * posterior_samples$y) / 
                 (1 + exp(posterior_samples$x + 20 * posterior_samples$y))
```

# 90% Credible Interval for \( p_{20} \)

```{r}
credible_interval <- quantile(p_success_20, c(0.05, 0.95))
credible_interval
```

# Summary of Results

- **Posterior Mean of Success Probability (ACT = 20):** `r mean(p_success_20)`
- **90% Credible Interval for Success Probability:** `r credible_interval`

